---
title: "Crushed Stone Aggregates Demand"
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
# Load the packages needed for this project.
# install.packages("corrplot")

library(dplyr)
library(ggplot2)
library(corrplot)

stone_data <- read.csv("/Users/jschnee12/Library/Mobile Documents/com~apple~CloudDocs/MDATA600-Crushed-Stone/data/processed/crushed_stone_state_year.csv")
```

## 1. Introduction

### Research questions

1.  Which construction, economic, capacity, and cost indicators best explain state-level crushed-stone production from 2015–2023?
2.  To what extent do prior-year state highway capital outlays predict state-level crushed-stone production across the United States from 2015–2023?
3.  How accurately can state-level construction GDP, highway capital outlays, and residential building permits predict crushed-stone production from 2015–2023?

### Project purpose

Crushed stone is a product found at the foundation of our economy. It is used in many segments of the economy, from residential housing, data center construction, road construction, and erosion control. Since this product is at the foundation of the economy, the demand for this material reveals the health of the broader economy.

This project is an explanatory project. The intention is to analyze which publically tracked economic indicators best explain changes in crushed stone demand.

The unit of observation in this project is one state-year combination per row. The full dataset includes 9 years of data, 2015-2023, for each of the 50 U.S. states. There are 450 rows and 80 columns.  

The outcome variable for each of the three research questions is `crushed_stone_tons`. This variable acts as a proxy for crushed stone demand since this measure is difficult to quantify outside of official crushed stone tons published by the United States Geological Survey. The main predictors vary across public sources and mainly represent crushed stone capacity, demand, economic variables, and input costs.

This analysis only includes state-level data from the United States. This project is explanatory in nature, so forecasting into future years is not advised. 

## 2. Data

### Data source and structure

- Report the state and year coverage.
- Identify any source limitations or withheld value.

The source file is a consolidation of multiple public data sources. Some metrics are posted throughout the year, so a few data points are averages or summations for the entire calendar year. The data was pulled from the following sources:

API-based sources

- **BEA Regional API:** construction GDP, total GDP, personal income.
- **BLS Public Data API:** construction employment series.
- **FRED API:** mortgage rates and CPI.

File exports/downloads

- **USGS Excel workbook:** crushed stone production, value, and unit value.
- **Census BPS text files:** building permits and housing units.
- **FHWA SF-4C tables/workbook:** highway capital outlays.
- **Census Construction Spending Excel files:** private and state/local construction spending.
- **EIA CSV and Excel codebook:** industrial energy prices.
- **MSHA ZIP archives:** mine and employment data.

```{r}
#| label: coverage
str(stone_data)
```

### Research Question 1 Dataset

For research question 1, I filtered down the dataset to only include 2015 - 2023, combined Census authorized building units into a summation, `residential_permit_units`, and selected the following columns:

- `state`
- `year`
- `crushed_stone_tons`
- `construction_real_gdp`
- `capital_outlay_total`
- `residential_permit_units`
- `private_nonresidential_spending_nominal_millions`
- `state_local_construction_spending_nominal_millions`
- `real_gdp`
- `personal_income`
- `construction_employment_thousands`
- `active_stone_mines`
- `stone_mine_employee_hours`
- `industrial_energy_price_per_mmbtu`

```{r}
rq1_data <- stone_data |>
  filter(year >= 2015, year <= 2023) |>
  mutate(
    residential_permit_units =
      units_1unit +
      units_2unit +
      units_34unit +
      units_5punit
  ) |>
  select(
    state,
    year,
    crushed_stone_tons,
    construction_real_gdp,
    capital_outlay_total,
    residential_permit_units,
    private_nonresidential_spending_nominal_millions,
    state_local_construction_spending_nominal_millions,
    real_gdp,
    personal_income,
    construction_employment_thousands,
    active_stone_mines,
    stone_mine_employee_hours,
    industrial_energy_price_per_mmbtu
  )

fields <- c("crushed_stone_tons",
    "construction_real_gdp",
    "capital_outlay_total",
    "residential_permit_units",
    "private_nonresidential_spending_nominal_millions",
    "state_local_construction_spending_nominal_millions",
    "real_gdp",
    "personal_income",
    "construction_employment_thousands",
    "active_stone_mines",
    "stone_mine_employee_hours",
    "industrial_energy_price_per_mmbtu"
    )
```

### Data-quality checks

```{r}
#| label: data-quality
rq1_data |>
  count(state, year, name = "n") |>
  filter(n > 1)

colSums(is.na(rq1_data))

rq1_data |>
  select(
    state,
    year,
    crushed_stone_tons,
    construction_employment_thousands
  ) |>
  filter(if_any(everything(), is.na))

rq1_data_clean <- rq1_data |>
  filter(!is.na(crushed_stone_tons))
```

There are 23 rows where the predictor is blank. I filtered these rows out of the dataset, so the remainder of the analysis will take place on `rq1_data_clean`.

## 3. Exploratory analysis

### Descriptive summary

```{r}
#| label: summary-table
summary(rq1_data_clean)
```

*Write a short interpretation of the table here.*

### Visualizing Scale of the Predictor

Explain what you want to learn about annual changes in production.

```{r}
#| label: annual-production
#| fig-width: 7
#| fig-height: 4.5
hist(rq1_data_clean$crushed_stone_tons / 1000000,
     breaks = 25,
     col = "steelblue",
     main = "Crushed Stone Production Distribution",
     xlab = "Production (million metric tons)")
```

*Write a short interpretation of the figure here.*

### Correlations

Describe the pairwise linear relationships among the numeric variables. Note any strong relationships among predictors that may affect the regression.

```{r}
#| label: correlations
# Open a larger plot window
dev.new(width = 14, height = 11)

plot_labels <- c(
  "Stone production",
  "Construction GDP",
  "Highway outlays",
  "Residential permits",
  "Private nonres. spend",
  "State/local const. spend",
  "Total real GDP",
  "Personal income",
  "Construction employment",
  "Active stone mines",
  "Mine employee hours",
  "Industrial energy price"
)

cor_matrix <- cor(
  rq1_data_clean[, fields],
  use = "complete.obs"
)

colnames(cor_matrix) <- plot_labels
rownames(cor_matrix) <- plot_labels

corrplot(
  cor_matrix,
  method = "color",
  type = "lower",
  order = "hclust",
  addCoef.col = "black",
  tl.col = "black",
  tl.srt = 35,
  tl.cex = 0.8,
  number.cex = 0.6,
  col = colorRampPalette(c("#B2182B", "white", "#2166AC"))(200),
  title = "Correlation Matrix: Crushed-Stone Production Indicators",
  mar = c(0, 0, 2, 0)
)
```

### Relationships with the predictor

```{r}
#| label: predictor-scatterplot
#| fig-width: 7
#| fig-height: 4.5
plot(
  rq1_data_clean$private_nonresidential_spending_nominal_millions,
  rq1_data_clean$crushed_stone_tons / 1000000,
  pch = 19,
  col = "pink",
  main = "Production vs Housing Units",
  xlab = "Residential Permit Units",
  ylab = "Production (million metric tons)"
) 
plot(
  rq1_data_clean$capital_outlay_total,
  rq1_data_clean$crushed_stone_tons / 1000000,
  pch = 19,
  col = "orange",
  main = "Production vs Contruction Spending",
  xlab = "Contruction Spending",
  ylab = "Production (million metric tons)"
) 
plot(
  rq1_data_clean$residential_permit_units,
  rq1_data_clean$crushed_stone_tons / 1000000,
  pch = 19,
  col = "purple",
  main = "Production vs Active Mines",
  xlab = "Active Mines",
  ylab = "Production (million metric tons)"
) 
plot(
  rq1_data_clean$construction_employment_thousands,
  rq1_data_clean$crushed_stone_tons / 1000000,
  pch = 19,
  col = "blue",
  main = "Production vs Construction Real GDP",
  xlab = "Construction Real GDP",
  ylab = "Production (million metric tons)"
)
```

## 4. Model

### Modeling strategy

- Define the baseline model.
- Define the full model.
- Explain why these predictors were selected.
- State how coefficients will be interpreted.
- State what the model does not establish.

```{r}
#| label: fit-models
# Fit a mean-only baseline model.
# Fit a multiple linear regression with the selected predictors.

# Simple regression
# Model 1: Private nonresidential spending
model1 <- lm(
  crushed_stone_tons ~ private_nonresidential_spending_nominal_millions,
  data = rq1_data_clean
)
summary(model1)

# Model 2: Capital outlay
model2 <- lm(
  crushed_stone_tons ~ capital_outlay_total,
  data = rq1_data_clean
)
summary(model2)

# Model 3: Residential permit units
model3 <- lm(
  crushed_stone_tons ~ residential_permit_units,
  data = rq1_data_clean
)
summary(model3)

# Model 4: Construction employment
model4 <- lm(
  crushed_stone_tons ~ construction_employment_thousands,
  data = rq1_data_clean
)
summary(model4)

# Multiple regression
model5 <- lm(
  crushed_stone_tons ~
    private_nonresidential_spending_nominal_millions +
    capital_outlay_total +
    residential_permit_units +
    construction_employment_thousands,
  data = rq1_data_clean
)

summary(model5)
```

### Coefficients

Report the coefficient estimates, standard errors, p-values, and confidence intervals.

```{r}
#| label: coefficients
# Tidy the full model and select the statistics you want to report.
```

*Interpret the direction, size, and uncertainty of the important coefficients in their original units.*

### Model fit

Compare the baseline and full models using measures such as:

- R-squared
- Adjusted R-squared
- In-sample RMSE

```{r}
#| label: model-fit
# Build a comparison table for the baseline and full models.
```

*Explain whether the full model improves in-sample fit and what that comparison does not tell you.*

## 5. Basic diagnostics

Use diagnostic plots or numerical checks to evaluate the main linear-model assumptions.

### Residuals versus fitted values

```{r}
#| label: residuals-fitted
#| fig-width: 7
#| fig-height: 4.5
# Create a residuals-versus-fitted plot.
```

*Comment on nonlinearity, unequal variance, outliers, or other visible patterns.*

### Normal Q-Q plot

```{r}
#| label: normal-qq
#| fig-width: 7
#| fig-height: 4.5
# Create a normal Q-Q plot of the model residuals.
```

*Comment on departures from normality, especially in the tails.*

### Diagnostic conclusion

State which assumptions appear reasonable, which appear questionable, and whether you will make any correction or leave the simple model unchanged.

## 6. Findings and limitations

### Main findings

Summarize the most important results in plain language:

- Which predictors have the strongest associations?
- What are the directions of the conditional coefficients?
- How much variation does the model explain?
- How does it compare with the baseline?

*Write your findings here. Distinguish clearly between association and causation.*

### Limitations

Address the limitations that apply to your analysis, such as:

- Short time span
- State-level aggregation
- Repeated observations of the same states
- Missing or withheld source values
- Measurement differences across sources
- Omitted variables and confounding
- Multicollinearity
- Diagnostic problems
- In-sample evaluation only

*Write your limitations here.*

### Conclusion

Return to the research question and state what the analysis supports, what it does not support, and what a stronger follow-up analysis could do.

*Write your conclusion here.*

## 7. AI-use disclosure

Write an accurate disclosure describing:

- Whether AI tools were used
- Which tasks they assisted with
- Which parts you reviewed, ran, changed, interpreted, or wrote yourself

> Replace this prompt with your own accurate disclosure before submitting.
